---
title: "A Gentrification Index for Chicago"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: flatly
---

```{r setup, include=FALSE}

# Packages & WD
library(tidyverse);library(sf);library(leaflet);library(leaflet.extras);
library(geojsonio);library(geojsonsf);library(hrbrthemes);library(ggthemes);library(DT);
library(shinyWidgets)
setwd(here::here()) # preserves relative paths on others systems as long as they open in fresh session

# Import data
vars <- read.csv("GI_Values_Tidy.csv") # GI data
vars$YEAR <- factor(vars$YEAR) # For grouping GI data by study period
# cca <- st_read("Shapefiles/CCA_GI.shp") # CCA shapefiles
# cca$CCA <- as.character(paste(cca$CCA))
# colnames(cca)[1] <- "CCA.NUM"
cca <- geojson_sf("CCA.geojson")

# Global lists
studyYears <- list(
  "1970" = "1970", 
  "1980" = "1980", 
  "1990" = "1990", 
  "2000" = "2000", 
  "2010" = "2010", 
  "2008 - 2012" = "2008 - 2012", 
  "2013 - 2017" = "2013 - 2017"
  )

studyVars <- list( 
  "Population" = "Population", 
  "Median Family Income" = "Median Family Income", 
  "Median Home value" = "Median Home Value", 
  "Percent White Population" = "Pct White", 
  "Percent Black Population" = "Pct Black", 
  "Percent Latinx Population" = "Pct Hispanic", 
  "Percent Children" = "Pct Children", 
  "Percent Elderly" = "Pct Elderly", 
  "Families in Poverty" = "Pct Families in Poverty", 
  "Female Headed Households with Children" = "Pct Female Headed Household with Kids", 
  "Owner-Occupied" = "Pct Owner Occupied", 
  "Managerial or Prof. Workers" = "Pct Managerial or Professional Workers", 
  "College Educated" = "Pct College Educated", 
  "Private School Attendance" = "Pct Private School Attendance"
  )

# Import functions
source("functions.R", local = TRUE) # required "local = TRUE" for shiny apps
# index_table creates a table of GI values for the specified year
# set_typology applies the GI types to each CCA
# change_table creates the table that is joined to the CCA shapefile for use with leaflet. It uses both index_table and set_typology.
# for gathering GI change, index, and population values simply run change_table(from, to)
# For displaying the underlying values of the GI variables you'll need to use the vars table. 
```


Sidebar {.sidebar}
======================================================================

```{r inputs}

shinyWidgets::sliderTextInput( # Slider input for study period selection
  inputId = "studyPeriod", 
  label = "Years:", 
  choices = studyYears,
  selected = c("1990", "2010"),
  grid = TRUE,
  from_max = "2008 - 2012",
  to_min = "1980"
  )

selectInput('studyVar', label = "Index Variable", choices = studyVars, selected = "Population") # Input for the study variable used in the Over Time tab

### Reactives
studyVar <- reactive ({
  input$studyVar
})

from <- reactive({ 
  input$studyPeriod[1]
})

to <- reactive({
  input$studyPeriod[2]
})

cca_gi <- reactive({ 
  change_table(from(), to()) %>% 
    left_join(cca, .) %>%
    select(-Cmmnt_N, -CCA_Cmb)
  })

```

Use the **sliders** and **dropdown menu** above to change the study period and focus on a specific variable in the **Over Time** tab.

**Click** on a community area on the map to see more information about it.

Use the **By Community Area** tab to examine gentrification index scores by Chicago Community Area.  

Use the **Over Time** tab to investigate trends by neighborhood or index component over time.  

To learn more about the project, click the **About** tab.  

Application authors:  
[Ethan Jantz](https://github.com/EthanJantz) and  
[Sean Connelly](https://sean-connelly.github.io/) 

[Voorhees Center](https://voorheescenter.uic.edu/)

By Community Area
======================================================================

Row
-----------------------------------------------------------------------

### Gentrification Index Map

```{r by CCA map}

output$map <- renderLeaflet({

  
  # Palette
  pal <- reactive({

   colorFactor(
     palette = c(
       "#FFFC86",
       "#FFFDB9",
       "#B8CCE4",
       "#4F81BC",
       "#D1FFC0",
       "#89CD66",
       "#D5CEB5",
       "#CDAA66",
       "#886F46"),
     levels = c(
       "1 - Upper Class",
       "2 - Middle Class",
       "3 - Poverty",
       "4 - Extreme Poverty",
       "5 - Not Gentrification",
       "6 - Gentrification", 
       "7 - Mild Decline",
       "8 - Moderate Decline",
       "9 - Serious Decline"
     ))
    
  })
  
  # Map
  map <- leaflet(cca_gi()) %>%
    addProviderTiles("CartoDB.Positron") %>%
    clearShapes() %>%
    addPolygons(stroke = TRUE, color = "#444444", weight = 1, smoothFactor = 0, 
                fillColor = ~pal()(cca_gi()$type), fillOpacity = 0.8, 
                layerId = ~CCA.NUM) %>%
    addLegend(position = "bottomleft", pal = pal(), 
              values = cca_gi()$type, title = "Gentrification Index<br>Value")
  
  map
  
})

# Click event for the map (will use to generate chart)
click_cca <- eventReactive(input$map_shape_click, {
  
  x <- input$map_shape_click
  y <- x$id
  return(y)
  
})

# Add the clicked CCA to the map in aqua, and remove when a new one is clicked
observe({
  
  req(click_cca()) # do this if click_cca() is not null
  
  map <- leafletProxy("map") %>%
    removeShape("highlight_cca") %>%
    addPolygons(data = cca_gi()[cca_gi()$CCA.NUM == click_cca(), ], fill = FALSE,
                color = "#FF9933", opacity = 1, layerId = "highlight_cca")
  
})

# Fetch data for the clicked CCA
selected_cca <- reactive({
  
  if(is.null(input$map_shape_click)){
  cca_gi()
  } else {
  cca_gi() %>% 
    filter(CCA.NUM == click_cca())
  }
})

leafletOutput("map")  

```

### Data, Selected Community Area (Click on the map to show table)

```{r DataTable output}

# Data table
output$cca_table <- DT::renderDataTable({
  
  DT::datatable(selected_cca() %>% 
                  st_drop_geometry() %>%
                  mutate(CCA.NUM = as.numeric(CCA.NUM)) %>%
                  arrange(CCA.NUM),
                extensions = "Buttons",
                options = list(dom = "Bfrtip",
                               buttons = c("copy", "csv", "excel", "pdf", "print"),
                               scrollX = TRUE,
                               scrollY = TRUE))
  
  })

DT::dataTableOutput("cca_table")

```



Over Time
======================================================================

Row
-----------------------------------------------------------------------

### Map

```{r over time map}
# This section will include a map similar to the one in the "By Community Area" section. 
# The click event will allow the user to select which CCA they want to view details on
# Alternatively, the user can select the CCA from a drop down menu
output$map2 <- renderLeaflet({

  
  # Palette
  pal <- reactive({
    
    # colorFactor(palette = c("#FEFD03", "#FFFFCD", "#B9CDE5", "#4F81BC", "#D2FFBE"),
    #             domain = cca_gi()$type)
   colorFactor(
     palette = c(
       "#FFFC86",
       "#FFFDB9",
       "#B8CCE4",
       "#4F81BC",
       "#D1FFC0",
       "#89CD66",
       "#D5CEB5",
       "#CDAA66",
       "#886F46"),
     levels = c(
       "1 - Upper Class",
       "2 - Middle Class",
       "3 - Poverty",
       "4 - Extreme Poverty",
       "5 - Not Gentrification",
       "6 - Gentrification", 
       "7 - Mild Decline",
       "8 - Moderate Decline",
       "9 - Serious Decline"
     ))
    
  })
  
  # Map
  map <- leaflet(cca_gi()) %>%
    addProviderTiles("CartoDB.Positron") %>%
    clearShapes() %>%
    addPolygons(stroke = TRUE, color = "#444444", weight = 1, smoothFactor = 0, 
                fillColor = ~pal()(cca_gi()$type), fillOpacity = 0.8, 
                layerId = ~COMMUNITY.AREA) %>%
    addLegend(position = "bottomleft", pal = pal(), 
              values = cca_gi()$type, title = "Gentrification Index<br>Value")
  
  map
  
})

# Click event for the map (will use to generate chart)
click_cca.ot <- eventReactive(input$map2_shape_click, {
  
  x <- input$map2_shape_click
  y <- x$id
  return(y)
  
})

# Add the clicked CCA to the map in aqua, and remove when a new one is clicked
observe({
  
  req(click_cca.ot()) # do this if click_cca() is not null
  
  map <- leafletProxy("map2") %>%
    removeShape("highlight_cca") %>%
    addPolygons(data = cca_gi()[cca_gi()$COMMUNITY.AREA == click_cca.ot(), ], fill = FALSE,
                color = "#FF9933", opacity = 1, layerId = "highlight_cca")
  
})

# Fetch data for the clicked CCA
selected_cca.ot <- reactive({
  
  if(is.null(input$map2_shape_click)){
  cca_gi()
  } else {
  cca_gi() %>% 
    filter(COMMUNITY.AREA == click_cca.ot())
  }
})

leafletOutput("map2")  
```


### Chart

```{r ot-chart}
# A ggplot line graph of selecetd index values over time

output$chart <- renderPlot({
  req(click_cca.ot())
  chart <- var_over_time(click_cca.ot(), from(), to(), studyVar())
  chart
})

plotOutput("chart")

```

Row
-----------------------------------------------------------------------

### Selected CCA Yearly Values

```{r ot-grid}
output$cca_val_byYear <- DT::renderDataTable({
  
  req(click_cca.ot())
  DT::datatable(vars %>%
    filter(COMMUNITY.AREA == click_cca.ot()) %>%
    pivot_wider(names_from = GI_VAR, values_from = VALUE) %>%
    mutate_if(is.numeric, round, 1), 
  extensions = "Buttons",
  options = list(dom = "Bfrtip",
  buttons = c("copy", "csv", "excel", "pdf", "print"),
  scrollX = TRUE,
  scrollY = TRUE))
})

DT::dataTableOutput("cca_val_byYear") 

### Future possibilities
# A table of the 14 GI components with icons showing their change during the study period
# Green-Red icon to indicate change along with the start and end values from the ot-chart
```


About
============================================================================

The Socioeconomic Change of Chicago’s Community Areas, also known as the Gentrification Index, examines neighborhood change across Chicago from 1970 to 2010. It uses key indicators to measure how much a neighborhood's wealth or poverty has changed in this time.

The [Socioeconomic Change of Chicago’s Community Areas (1970-2010)](http://voorheescenter.red.uic.edu/wp-content/uploads/sites/122/2017/10/Voorhees-Center-Gentrification-Index-Oct-14.pdf) shows that inequality is growing in Chicago. Some neighborhoods have grown wealthier, while others have grown poorer. At the same time, the number of middle-class neighborhoods has gone down.

An Appendix, showing the detailed information for each community area, is also [available](http://voorheescenter.red.uic.edu/wp-content/uploads/sites/122/2017/10/Appendix-Oct-14.pdf).

For information about how to mitigate the impact of gentrification, check out the [toolkit](http://voorheescenter.uic.edu/wp-content/uploads/sites/122/2017/10/Gentrification-and-Neighborhood-Change-Toolkit.pdf).

To see presentations for the 2015 Voorhees Center symposium called *Taking Control of our Neighborhoods’ Future*, look [here](https://voorheescenter.wordpress.com/2015/05/27/investment-without-displacement-communities-gather-to-discuss-gentrification-at-the-voorhees-center-symposium/).

Questions? Try the [Gentrification FAQ](http://voorheescenter.uic.edu/what-we-do/areas-of-research/gentrification-index/gentrification-faq/).

For more, see WBEZ’s interactive website on gentrification [here](http://interactive.wbez.org/gentrification/).

*Report Summary*: [English](http://voorheescenter.uic.edu/wp-content/uploads/sites/122/2018/03/GI-One-Pager.pdf), [Español](http://voorheescenter.uic.edu/wp-content/uploads/sites/122/2018/03/GI-Una-Pagina1.pdf)
